---
title: "More Graphs"
---

# More Graphs

```{r setup-more-graphs}
library(readr)
library(dplyr)
library(ggplot2)
library(broom)
library(knitr)
library(tidyr)

theme_set(theme_minimal(base_size = 14))

# Read the dataset
df <- read_tsv(
  "PLoSOneData_ExamsDisWomen.csv",
  na = c("", " "),
  locale = locale(encoding = "UTF-8"),
  guess_max = 10000
) %>%
  mutate(
    Gender = factor(Gender, levels = c(0, 1), labels = c("Women", "Men")),
    URM = factor(URM, levels = c(0, 1), labels = c("Non-URM", "URM")),
    YearSchool = factor(YearSchool),
    GRADING_BASIS_ENRL = factor(GRADING_BASIS_ENRL),
    SUBJECT = factor(SUBJECT),
    AGE = as.numeric(AGE),
    Coursecode = as.character(Coursecode)
  )
```

The age-based plot illustrates how exam percentages vary by gender and whether that relationship differs for URM and non-URM students.

```{r more-graphs-model2}
ggplot(df, aes(x = AGE, y = Exams_percent, color = Gender)) +
  geom_jitter(alpha = 0.45, width = 0.2, height = 0.01, size = 1.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  facet_wrap(~ URM) +
  labs(
    title = "Exam performance by age, gender, and URM status",
    x = "Age",
    y = "Exam percentage",
    color = "Gender"
  ) +
  scale_color_brewer(palette = "Dark2")
```

The boxplot summarizes how exam percentages are distributed across gender within each grading scheme, making it easier to compare the spread of outcomes by course structure.

```{r more-graphs-model3}
ggplot(df, aes(x = Gender, y = Exams_percent, fill = Gender)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.2) +
  facet_wrap(~ GRADING_BASIS_ENRL) +
  labs(
    title = "Exam performance by gender across grading schemes",
    x = "Gender",
    y = "Exam percentage"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme(legend.position = "none")
```

The histogram below shows the distribution of exam percentages across students, which helps reveal whether exam scores are clustered around particular values or broadly dispersed.

```{r more-graphs-exam-histogram}
ggplot(df, aes(x = Exams_percent)) +
  geom_histogram(bins = 25, fill = "#4C78A8", color = "white") +
  labs(
    title = "Distribution of exam percentages",
    x = "Exam percentage",
    y = "Count"
  )
```

The bar chart below displays the count of courses by grading scheme, illustrating which grading bases are most common in the dataset.

```{r more-graphs-grading-chart}
ggplot(df, aes(x = GRADING_BASIS_ENRL, fill = GRADING_BASIS_ENRL)) +
  geom_bar() +
  labs(
    title = "Distribution of grading schemes",
    x = "Grading basis",
    y = "Count"
  ) +
  scale_fill_brewer(palette = "Dark2") +
  theme(legend.position = "none")
```
