---
title: "Preceptor Table Model"
format:
  html:
    toc: true
    code-fold: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(readr)
library(dplyr)
library(knitr)
library(tidyr)

# Read and prepare the dataset for the preceptor-style table.
df <- read_tsv(
  "PLoSOneData_ExamsDisWomen.csv",
  na = c("", " "),
  locale = locale(encoding = "UTF-8"),
  guess_max = 10000
) %>%
  mutate(
    Gender = factor(Gender, levels = c(0, 1), labels = c("Women", "Men")),
    URM = factor(URM, levels = c(0, 1), labels = c("Non-URM", "URM")),
    YearSchool = factor(YearSchool),
    GRADING_BASIS_ENRL = factor(GRADING_BASIS_ENRL),
    SUBJECT = factor(SUBJECT),
    AGE = as.numeric(AGE)
  )
```

This page presents a compact preceptor-style summary of the dataset. It includes both a preview of the observations and a variable-level summary drawn from the actual study data.

## Data-generating mechanism

The model on this page uses a linear regression framework. In this setting, the exam percentage outcome is treated as being generated by a combination of predictors such as gender, ACT score, age, URM status, class standing, and grading basis, with the relationship described by a linear model.

$$
\begin{aligned}
\text{Exams\_percent}_i
&= \beta_0
+ \beta_1(\text{AGE}_i)
+ \beta_2(\text{Gender}_{\text{Men},i})
+ \beta_3(\text{URM}_i)
+ \beta_4(\text{COMP\_ACT\_SCORE}_i) \\
&\quad
+ \beta_5(\text{YearSchool}_{\text{Sophomore},i})
+ \beta_6(\text{YearSchool}_{\text{Junior},i})
+ \beta_7(\text{YearSchool}_{\text{Senior},i})
+ \varepsilon_i,
\end{aligned}
$$

where

$$
\varepsilon_i \sim N(0,\sigma^2).
$$

```{r preview-table}
preview_tbl <- df %>%
  select(
    SUBJECT,
    Coursecode,
    GRADING_BASIS_ENRL,
    Gender,
    AGE,
    YearSchool,
    COMP_ACT_SCORE,
    URM,
    Exams_percent,
    Course_percent,
    NonExam_percent
  ) %>%
  slice_head(n = 20)

kable(preview_tbl, digits = 3, caption = "Preview of the study dataset")
```

```{r summary-table}
summary_tbl <- tibble::tibble(
  Variable = c(
    "SUBJECT",
    "Coursecode",
    "GRADING_BASIS_ENRL",
    "Gender",
    "AGE",
    "YearSchool",
    "COMP_ACT_SCORE",
    "URM",
    "Exams_percent",
    "Course_percent",
    "NonExam_percent"
  ),
  Type = c(
    "Categorical",
    "Categorical",
    "Categorical",
    "Categorical",
    "Numeric",
    "Categorical",
    "Numeric",
    "Categorical",
    "Numeric",
    "Numeric",
    "Numeric"
  ),
  Missing = c(
    sum(is.na(df$SUBJECT)),
    sum(is.na(df$Coursecode)),
    sum(is.na(df$GRADING_BASIS_ENRL)),
    sum(is.na(df$Gender)),
    sum(is.na(df$AGE)),
    sum(is.na(df$YearSchool)),
    sum(is.na(df$COMP_ACT_SCORE)),
    sum(is.na(df$URM)),
    sum(is.na(df$Exams_percent)),
    sum(is.na(df$Course_percent)),
    sum(is.na(df$NonExam_percent))
  ),
  `Sample values` = c(
    paste(unique(na.omit(df$SUBJECT))[1:min(3, n_distinct(na.omit(df$SUBJECT)))], collapse = ", "),
    paste(unique(na.omit(df$Coursecode))[1:min(3, n_distinct(na.omit(df$Coursecode)))], collapse = ", "),
    paste(unique(na.omit(df$GRADING_BASIS_ENRL))[1:min(3, n_distinct(na.omit(df$GRADING_BASIS_ENRL)))], collapse = ", "),
    paste(unique(na.omit(df$Gender))[1:min(3, n_distinct(na.omit(df$Gender)))], collapse = ", "),
    paste(round(quantile(df$AGE, probs = c(0.25, 0.5, 0.75), na.rm = TRUE), 1), collapse = ", "),
    paste(unique(na.omit(df$YearSchool))[1:min(3, n_distinct(na.omit(df$YearSchool)))], collapse = ", "),
    paste(round(quantile(df$COMP_ACT_SCORE, probs = c(0.25, 0.5, 0.75), na.rm = TRUE), 1), collapse = ", "),
    paste(unique(na.omit(df$URM))[1:min(3, n_distinct(na.omit(df$URM)))], collapse = ", "),
    paste(round(quantile(df$Exams_percent, probs = c(0.25, 0.5, 0.75), na.rm = TRUE), 3), collapse = ", "),
    paste(round(quantile(df$Course_percent, probs = c(0.25, 0.5, 0.75), na.rm = TRUE), 3), collapse = ", "),
    paste(round(quantile(df$NonExam_percent, probs = c(0.25, 0.5, 0.75), na.rm = TRUE), 3), collapse = ", ")
  )
)

kable(summary_tbl, caption = "Variable summary for the dataset")
```
