---
title: "Gender, assessment, and performance in introductory biology"
author: "Ilvika Ravat"
format:
  html:
    toc: true
    theme: cosmo
    code-fold: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(readr)
library(dplyr)
library(ggplot2)
library(broom)
library(knitr)
library(tidyr)

theme_set(theme_minimal(base_size = 14))

# Read the tab-delimited dataset and clean the main grouping variables.
df <- read_tsv(
  "PLoSOneData_ExamsDisWomen.csv",
  na = c("", " "),
  locale = locale(encoding = "UTF-8"),
  guess_max = 10000
) %>%
  mutate(
    Gender = factor(Gender, levels = c(0, 1), labels = c("Women", "Men")),
    URM = factor(URM, levels = c(0, 1), labels = c("Non-URM", "URM")),
    YearSchool = factor(YearSchool),
    GRADING_BASIS_ENRL = factor(GRADING_BASIS_ENRL),
    SUBJECT = factor(SUBJECT),
    AGE = as.numeric(AGE),
    Coursecode = as.character(Coursecode)
  )
```

# Overview {.panel-tabset}

## Study overview

This report explores how exam-based assessment relates to student performance in introductory biology using the data file provided in this project. The analysis focuses on how student outcomes vary by gender, academic preparation, underrepresented minority status, course grading design, and other contextual factors.

```{r overview-table}
overview_tbl <- tibble::tibble(
  Metric = c("Students", "Courses", "Gender groups", "URM groups", "Grading schemes"),
  Value = c(nrow(df), n_distinct(df$Coursecode), n_distinct(df$Gender), n_distinct(df$URM), n_distinct(df$GRADING_BASIS_ENRL))
)

kable(overview_tbl, caption = "Dataset overview")
```

## Variable dictionary

The table below explains the variables in the dataset. The response variables describe the share of course performance associated with exams, coursework, and non-exam assessment, while the predictor variables capture student background and course context.

```{r variable-table}
var_desc <- tibble::tribble(
  ~Variable, ~Type, ~Description,
  "SUBJECT", "Categorical", "The academic subject of the course, such as introductory biology.",
  "Coursecode", "Categorical", "The course identifier used to distinguish specific sections or offerings.",
  "GRADING_BASIS_ENRL", "Categorical", "The grading basis used for enrollment in the course, such as A-F or S-N.",
  "Gender", "Categorical", "A binary indicator for student gender, recoded here as Women and Men.",
  "AGE", "Numeric", "Student age in years.",
  "YearSchool", "Categorical", "The student's class standing, such as freshman, sophomore, junior, or senior.",
  "COMP_ACT_SCORE", "Numeric", "Composite ACT score, used as a proxy for academic preparation.",
  "URM", "Categorical", "Underrepresented minority indicator, recoded here as URM and Non-URM.",
  "msl01", "Numeric", "One item from the motivation and learning-strategy survey instrument.",
  "msl02", "Numeric", "One item from the motivation and learning-strategy survey instrument.",
  "msl03", "Numeric", "One item from the motivation and learning-strategy survey instrument.",
  "msl04", "Numeric", "One item from the motivation and learning-strategy survey instrument.",
  "msl05", "Numeric", "One item from the motivation and learning-strategy survey instrument.",
  "msl06", "Numeric", "One item from the motivation and learning-strategy survey instrument.",
  "msl07", "Numeric", "One item from the motivation and learning-strategy survey instrument.",
  "msl08", "Numeric", "One item from the motivation and learning-strategy survey instrument.",
  "msl09", "Numeric", "One item from the motivation and learning-strategy survey instrument.",
  "msl10", "Numeric", "One item from the motivation and learning-strategy survey instrument.",
  "msl11", "Numeric", "One item from the motivation and learning-strategy survey instrument.",
  "msl12", "Numeric", "One item from the motivation and learning-strategy survey instrument.",
  "msl13", "Numeric", "One item from the motivation and learning-strategy survey instrument.",
  "msl14", "Numeric", "One item from the motivation and learning-strategy survey instrument.",
  "msl15", "Numeric", "One item from the motivation and learning-strategy survey instrument.",
  "msl16", "Numeric", "One item from the motivation and learning-strategy survey instrument.",
  "msl17", "Numeric", "One item from the motivation and learning-strategy survey instrument.",
  "msl18", "Numeric", "One item from the motivation and learning-strategy survey instrument.",
  "msl19", "Numeric", "One item from the motivation and learning-strategy survey instrument.",
  "msl20", "Numeric", "One item from the motivation and learning-strategy survey instrument.",
  "Exams_percent", "Numeric", "The share of the course grade associated with exams.",
  "Course_percent", "Numeric", "The share of the course grade associated with coursework.",
  "NonExam_percent", "Numeric", "The share of the course grade associated with non-exam activities.",
  "ZExams_percent", "Numeric", "Z-scored version of Exams_percent for standardized comparison.",
  "ZCourse_percent", "Numeric", "Z-scored version of Course_percent.",
  "ZNonExam_percent", "Numeric", "Z-scored version of NonExam_percent."
)

kable(var_desc, caption = "Variable descriptions")
```

## Results {.panel-tabset}

### Model 1: academic preparation and demographics

This first model examines whether exam performance varies with gender, age, ACT score, URM status, and class standing.

```{r model1}
mod1 <- lm(Exams_percent ~ Gender + AGE + COMP_ACT_SCORE + URM + YearSchool, data = df)

model1_tidy <- tidy(mod1, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(term = gsub("Gender", "Gender: ", term),
         term = gsub("URM", "URM: ", term))

kable(model1_tidy %>% select(term, estimate, std.error, p.value), digits = 3, caption = "Model 1 coefficients")
```

```{r model1-plot}
ggplot(df, aes(x = COMP_ACT_SCORE, y = Exams_percent, color = Gender, shape = URM)) +
  geom_point(alpha = 0.65, size = 2.2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  facet_wrap(~ YearSchool) +
  labs(
    title = "Exam performance by academic preparation and student background",
    x = "Composite ACT score",
    y = "Exam percentage",
    color = "Gender",
    shape = "URM status"
  ) +
  theme(legend.position = "bottom") +
  scale_color_brewer(palette = "Set2")
```

### Model 2: interaction between gender and URM status

This model tests whether the relationship between gender and exam performance changes by URM status.

```{r model2}
mod2 <- lm(Exams_percent ~ Gender * URM + AGE + COMP_ACT_SCORE + YearSchool, data = df)

model2_tidy <- tidy(mod2, conf.int = TRUE) %>%
  filter(term != "(Intercept)")

kable(model2_tidy %>% select(term, estimate, std.error, p.value), digits = 3, caption = "Model 2 coefficients")
```

```{r model2-plot}
ggplot(df, aes(x = AGE, y = Exams_percent, color = Gender)) +
  geom_jitter(alpha = 0.45, width = 0.2, height = 0.01, size = 1.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  facet_wrap(~ URM) +
  labs(
    title = "Exam performance by age, gender, and URM status",
    x = "Age",
    y = "Exam percentage",
    color = "Gender"
  ) +
  scale_color_brewer(palette = "Dark2")
```

### Model 3: grading basis and gender

This model asks whether the grading structure of the course changes the observed gender pattern.

```{r model3}
mod3 <- lm(Exams_percent ~ Gender * GRADING_BASIS_ENRL + AGE + COMP_ACT_SCORE + URM, data = df)

model3_tidy <- tidy(mod3, conf.int = TRUE) %>%
  filter(term != "(Intercept)")

kable(model3_tidy %>% select(term, estimate, std.error, p.value), digits = 3, caption = "Model 3 coefficients")
```

```{r model3-plot}
ggplot(df, aes(x = Gender, y = Exams_percent, fill = Gender)) +
  geom_boxplot(alpha = 0.8, outlier.alpha = 0.2) +
  facet_wrap(~ GRADING_BASIS_ENRL) +
  labs(
    title = "Exam performance by gender across grading schemes",
    x = "Gender",
    y = "Exam percentage"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme(legend.position = "none")
```

### Model 4: full contextual model

The final model combines the main student-background and course-context variables together.

```{r model4}
mod4 <- lm(Exams_percent ~ Gender + AGE + COMP_ACT_SCORE + URM + YearSchool + GRADING_BASIS_ENRL, data = df)

model4_tidy <- tidy(mod4, conf.int = TRUE) %>%
  filter(term != "(Intercept)")

kable(model4_tidy %>% select(term, estimate, std.error, p.value), digits = 3, caption = "Model 4 coefficients")
```

```{r model-comparison}
model_compare <- bind_rows(
  glance(mod1) %>% mutate(Model = "Model 1: demographics"),
  glance(mod2) %>% mutate(Model = "Model 2: gender × URM"),
  glance(mod3) %>% mutate(Model = "Model 3: grading basis"),
  glance(mod4) %>% mutate(Model = "Model 4: full context")
) %>%
  select(Model, r.squared, adj.r.squared, AIC)

kable(model_compare, digits = 3, caption = "Model comparison")
```

### Interpretation

Across these models, the pattern suggests that exam performance is shaped by both student characteristics and course design. Gender differences appear alongside preparation measures and grading context, which is why it is useful to compare multiple regression specifications rather than relying on a single model.

## Study context

The original study behind this dataset is available from Dryad here: <https://datadryad.org/dataset/doi%3A10.5061/dryad.53t31?utm>.

The study examines whether exam-based assessment may disadvantage women in introductory biology courses and provides the broader scientific context for the variables used in this analysis.

Data from: Exams disadvantage women in introductory biology
Cotner, Sehoya[1](https://datadryad.org/dataset/doi%3A10.5061/dryad.53t31?utm#aff239); Salehi, Shima[2](https://datadryad.org/dataset/doi%3A10.5061/dryad.53t31?utm#aff214); Ballen, Cissy J.[1](https://datadryad.org/dataset/doi%3A10.5061/dryad.53t31?utm#aff239)

Author affiliations

Published Oct 05, 2018 on Dryad. https://doi.org/10.5061/dryad.53t31            

Abstract

The gender gap in STEM fields has prompted a great deal of discussion, but what factors underlie performance deficits remain poorly understood. We show that female students underperformed on exams compared to their male counterparts across ten large introductory biology course sections in fall 2016 (N > 1500 students). Females also reported higher levels of test anxiety and course-relevant science interest. Results from mediation analyses revealed an intriguing pattern: for female students only, and regardless of their academic standing, test anxiety negatively impacted exam performance, while interest in the course-specific science topics increased exam performance. Thus, instructors seeking equitable classrooms can aim to decrease test anxiety and increase student interest in science course content. We provide strategies for mitigating test anxiety and suggestions for alignment of course content with student interest, with the hope of successfully reimagining the STEM pathway as one that is equally accessible to all.
